\pt{Acknowledgments}

\noindent{}I thank my advisors. I thank Professor Fred Fontaine, for keeping me on track and supporting me in my efforts to do research even as I diverge from EE into theoretical CS. Professor Fontaine has also been extremely generous in assisting me as my academic advisor of four years and in inviting me as an instructor for the ECE-210 course, all while serving as the Chair of Electrical Engineering and encouraging all my peers in the EE major to excel. I thank Professor Robert Marano, for sharing the passion in software engineering research and helping guide me towards research in programming education. I thank Professor Cyrus Omar at the University of Michigan, as well as the other members of the FPLab, for providing the technical mentorship for this project and allowing me access to your time and advice, despite the unconventional collaboration. The openness for an outsider such as myself was pivotal in allowing me to learn and accomplish as much as I was able to for this project.

I thank my peers in electrical engineering, for supporting me all this way. Among many peers, I thank the members of the 10B organization, for being my closest friends since the beginning of our shared Cooper experience and for the continual and occasionally flamboyantly-terrorizing level of support. I also thank Victor Zhang and Derek Lee for sharing the experience of various independent studies, senior projects, programming competitions, job applications, and the dual-degree program, all over the course of the past year.

Lastly, and most importantly, I thank my family, who have allowed me to spend all my efforts in school unhindered for the greater part of the past eight years, even through the panic of a pandemic. I thank my parents for the immense support in this process, and my sisters Jessica, Juliet, Josie, Sharon, and Jane for their emotional support. I thank the numerous family members that have allowed me to reside with them during my stay in the city: my grandma, Uncle Frank, Ben, Uncle John, and Aunt Amy. I hope I can repay your generosity someday.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
